Rectangulotopes

### 2024-04-22 draft/rectangulotopes.tex
getting there

We complete each node of~$S(R)$ and~$T(R)$ with a vertical (resp.~horizontal) leaf if it has no vertical (resp.~horizontal) child.
An example is given in~\cref{fig:strongRectangulation}.
If the rectangulation $R$ is diagonal, we retrieve the well-studied pair of twin binary trees~\cite{MR1417289,MR2914637}, see~\cref{fig:weakRectangulation}.
Recall that the \defn{inorder labeling} of a binary tree~$T$ is the labeling of the nodes of~$T$ such that the label of each node~$t$ of~$T$ is larger than all labels in the left subtree of~$t$ and larger than all labels in the right subtree of~$t$.
%\begin{lemma}
For any rectangle~$r$ of~$R$, the inorder label of the source~$s(r)$ in~$S(R)$ coincides with the inorder label of the target~$t(r)$ in~$T(R)$.
%\end{lemma}
This enables to unambiguously label the rectangles of~$R$ by the inorder.
The resulting labeling coincides with the NW--SE labeling of~\cite{ACFF24}.
From now on, the labels of the rectangles in a rectangulation are the inorder labels, and the two trees $T(R)$ and $S(R)$ are defined on the vertex set $[n]$ accordingly.
For each $i\in [n]$, we refer to its two subtrees in $T(R)$ as the \defn{horizontal} and \defn{vertical subtrees}, and similarly for $S(R)$.

@@

We denote by $i\verticalPattern j$ the situation where 
and by $i\horizontalPattern j$ the situation where \Jean{Complete with the right definition here.}

@@

\begin{theorem}[\cite{MR2871762}]
  The subsets $\mathcal{L}(P_w(R))$ of $\f{S}_n$, for all rectangulations $R$ of size~$n$, 
  are equivalence classes of a congruence relation $\weakeq$ on the weak Bruhat order on $\f{S}_n$.
  Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with weak rectangulations,  
  \item the minimal elements of the congruence classes are the twisted Baxter permutations,
  \item the maximal elements of the congruence classes are the co-twisted Baxter permutations,
  \item the congruence classes each contain a single Baxter permutation, and are one-to-one with Baxter permutations.
  \end{itemize}
\end{theorem}

@@

\vincent{Add here sentence about geometric interpretation of weak poset. Bijection cover relations weak poset -- facets -- flips...}

@@

\begin{theorem}[\cite{MR2864445,ACFF24}]
  The subsets $\mathcal{L}(P_s(R))$ of $\f{S}_n$, for all rectangulations $R$ of size~$n$, are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of the congruence classes are the 2-clumped permutations,
  \item the maximal elements of the congruence classes are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}

@@

@@

\cref{lem:yincount} essentially gives us the coefficient $\yin{w}^R_{i,j}$ in \cref{thm:strongRectangulotope}, except that the condition $\neg i\verticalPattern j$ is replaced by $i\succ_s j$.

Recall that $i\verticalPattern j$ denote the situation where the rectangles $r_i$ and $r_j$ of inorder labels $i$ and $j$ are such that $r_i$ and $r_j$ share a vertical segment, $r_i$ is on the left of $r_j$, and the bottom edge of $r_i$ is lower than the top edge of $r_j$.

Our aim is to provide a formula for the coordinates of the vertices of the strong rectangulotope that does not involve the computation of the strong poset or one of its linear extension.
For that purpose, we prove that for every pair $i<j\in [n]$ such that $i\prec_s j$, hence which do not have any associated yin arc, we have either $cv^{T,S}_{i,j} = 0$ (hence the product in \cref{lem:yincount} is 0 as well and adding it does not harm the result), or $i\verticalPattern j$.

\begin{lemma}
  \label{lem:whentosum}
  Let $R$ be a rectangulation of size~$n$, let $S$ and $T$ be its source and target trees, and let~$([n],\prec_s)$ be the associated strong order.
  Given a pair $i,j\in [n]$ such that $i<j$ and $i\prec_s j$, we have
  \[
    cv^{T,S}_{i,j} > 0 \quad \Longrightarrow \quad i \verticalPattern j.
  \]
\end{lemma}
\begin{proof}
  Consider the two rectangles $r_i$ and $r_j$ with inorder labels $i$ and $j$ in $R$.
  We first observe that the source vertex $s(r_j)$ is not contained in the interior of the lower left quadrant of origin $t(r_i)$.
  Indeed, if we suppose otherwise, then the line segment from $s(r_j)$ to $t(r_i)$ has positive slope, which implies that $j\tri i$, which in turn implies $j\prec_s i$, contradicting our hypothesis.

  We now claim that there can only be one segment on which leaves common to the vertical subtree of $i$ in $T$ and the vertical subtree of $j$ in $S$ may lie, and this segment is the vertical segment supporting the right edge of $r_i$ and the left edge of $r_j$.
  \Jean{This requires a case analysis. Any shortcut?}

  Let us assume that this segment is shared.
  If the bottom edge of $r_i$ is higher than the top edge of $r_j$, we must have $j\btri i$, which implies $j\prec_s i$, contradicting our hypothesis again.
  Hence we must have that the bottom edge of $r_i$ is lower than the top edge of $r_j$, hence that $i\verticalPattern j$, as claimed.
\end{proof}

We omit the proofs of the analogues of \cref{lem:yincount} and \cref{lem:whentosum} for the case of the yang arcs: The arguments are completely symmetric.
This concludes the proof of \cref{thm:strongRectangulotope}.

>>>>

We complete each node of~$S(R)$ and~$T(R)$ with a vertical (resp.~horizontal) leaf if it has no vertical (resp.~horizontal) child.
An example is given in~\cref{fig:strongRectangulation}.
If the rectangulation $R$ is diagonal, we retrieve the well-studied pair of twin binary trees~\cite{MR1417289,MR2914637}, see~\cref{fig:weakRectangulation}.

Recall that the \defn{inorder labeling} of a binary tree~$T$ is the labeling of the nodes of~$T$ such that the label of each node~$t$ of~$T$ is larger than all labels in the left subtree of~$t$ and larger than all labels in the right subtree of~$t$.
For any rectangle~$r$ of~$R$, the inorder label of the source~$s(r)$ in~$S(R)$ coincides with the inorder label of the target~$t(r)$ in~$T(R)$.
The inorder can also be retrieved from the above-below and left-right relationships defined before, as the transitive closure of the union of the below and right partial orders.
This enables to unambiguously label the rectangles of~$R$ by the inorder.
The resulting labeling coincides with the NW--SE labeling of~\cite{ACFF24}.
From now on, the labels of the rectangles in a rectangulation are the inorder labels, and the two trees $T(R)$ and $S(R)$ are defined on the vertex set $[n]$ accordingly.
For each $i\in [n]$, we refer to its two subtrees in $T(R)$ as the \defn{horizontal} and \defn{vertical subtrees}, and similarly for $S(R)$.

@@

We denote by $i\verticalPattern j$ the situation where the rectangles $r_i$ and $r_j$ of inorder labels $i$ and $j$ share a vertical segment, $r_i$ is on the left of $r_j$, and the bottom edge of $r_i$ is lower than the top edge of $r_j$.
and by $i\horizontalPattern j$ the situation where \Jean{Complete with the right definition here.}

@@

\begin{theorem}[\cite{MR2871762}]
  \label{thm:weakCong}
  The subsets $\mathcal{L}(P_w(R))$ of $\f{S}_n$, for all rectangulations $R$ of size~$n$, 
  are equivalence classes of a congruence relation $\weakeq$ on the weak Bruhat order on $\f{S}_n$.
  Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with weak rectangulations,  
  \item the minimal elements of the congruence classes are the twisted Baxter permutations,
  \item the maximal elements of the congruence classes are the co-twisted Baxter permutations,
  \item the congruence classes each contain a single Baxter permutation, and are one-to-one with Baxter permutations.
  \end{itemize}
\end{theorem}

@@

From \cref{thm:weakCong}, we obtain a bijection between the cover relations in the weak poset and the facets of the dual cone corresponding to the rectangulation, which are in turn one-to-one with feasible flips in the rectangulation.

@@

\begin{theorem}[\cite{MR2864445,ACFF24}]
  \label{thm:strongCong}
  The subsets $\mathcal{L}(P_s(R))$ of $\f{S}_n$, for all rectangulations $R$ of size~$n$, are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of the congruence classes are the 2-clumped permutations,
  \item the maximal elements of the congruence classes are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}

@@

Note that, similarly to the weak case, \cref{thm:strongCong} yields a bijection between the cover relations in the strong poset and the facets of the dual cone corresponding to the rectangulation, which are in turn one-to-one with feasible flips in the rectangulation.

@@

\cref{lem:yincount} essentially gives us the coefficient $\yin{w}^R_{i,j}$ in \cref{thm:strongRectangulotope}, except that the condition $\neg i\verticalPattern j$ is replaced by $i\succ_s j$.
Recall that $i\verticalPattern j$ denote the situation where the rectangles $r_i$ and $r_j$ of inorder labels $i$ and $j$ share a vertical segment, $r_i$ is on the left of $r_j$, and the bottom edge of $r_i$ is lower than the top edge of $r_j$.
Our aim is to provide a formula for the coordinates of the vertices of the strong rectangulotope that does not involve the computation of the strong poset or one of its linear extension.
For that purpose, we prove that for every pair $i<j\in [n]$ such that $i\prec_s j$, which do not have any associated yin arc, we have either $cv^{T,S}_{i,j} = 0$ (hence the product in \cref{lem:yincount} is 0 as well and adding it does not harm the result), or $i\verticalPattern j$.

\begin{lemma}
  \label{lem:whentosum}
  Let $R$ be a rectangulation of size~$n$, let $S$ and $T$ be its source and target trees, and let~$([n],\prec_s)$ be the associated strong order.
  Given a pair $i,j\in [n]$ such that $i<j$ and $i\prec_s j$, we have
  \[
    cv^{T,S}_{i,j} > 0 \quad \Longrightarrow \quad i \verticalPattern j.
  \]
\end{lemma}
\begin{proof}
  Consider the two rectangles $r_i$ and $r_j$ with inorder labels $i$ and $j$ in $R$.
  We first observe that the source vertex $s(r_j)$ is not contained in the interior of the lower left quadrant of origin $t(r_i)$.
  Indeed, if we suppose otherwise, then the line segment from $s(r_j)$ to $t(r_i)$ has positive slope, which implies that $j\tri i$, which in turn implies $j\prec_s i$, contradicting our hypothesis.
  We also observe that since $i < j$, by definition of the inorder, the rectangle $r_j$ must be below or on the right of $r_i$.
  Combining those two observations, we conclude that the left edge of $r_j$ cannot be on the left of the right edge of $r_i$.

  Therefore, there can only be one segment on which a leaf common to the vertical subtree of $i$ in $T$ and the vertical subtree of $j$ in $S$ may lie, which is the vertical segment supporting the right edge of $r_i$ and the left edge of $r_j$.
  Let us assume that this segment is shared by those two edges.
  If the bottom edge of $r_i$ is higher than the top edge of $r_j$, we must have $j\btri i$, which implies $j\prec_s i$, contradicting our hypothesis again.
  Thus the bottom edge of $r_i$ is lower than the top edge of $r_j$, and $i\verticalPattern j$, as claimed.
\end{proof}

We omit the proofs of the analogues of \cref{lem:yincount} and \cref{lem:whentosum} for the case of the yang arcs, which use completely symmetric arguments.
This concludes the proof of \cref{thm:strongRectangulotope}.
